Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-17 manuscript.tex
4.2

%: >>>>>>>> Laurent is here <<<<<<<<<<<<<<
\subsection{Scaling up to very large scale data}

\href{https://nbviewer.org/github/MouseLand/rastermap/blob/master/tutorial/tutorial.ipynb}{Rastermap} re-arranges neurons in the raster plot based on similarity of activity

\begin{itemize}
 \item
  https://www.janelia.org/lab/stringer-lab
\item
  described in~\citep{pachitariu_robustness_2018}
\item
  \href{https://github.com/MouseLand/rastermap}{rastermap}
\item
  deconvolution strategy
\item
  based on a linear model
\end{itemize}

\subsubsection{Stringer et al 2019, Nature~\citep{stringer_high-dimensional_2019}}\label{stringer-et-al-2019-nature-stringer2019nature}
\begin{itemize}
 \item
  ``A neuronal population encodes information most efficiently when its stimulus responses are high-dimensional and uncorrelated, and most robustly when they are lower-dimensional and correlated. Here we analysed the dimensionality of the encoding of natural images by large populations of neurons in the visual cortex of awake mice.''
\item
  Data availability: All of the processed deconvolved calcium traces are available on \href{https://figshare.com/articles/Recordings_of_ten_thousand_neurons_in_visual_cortex_in_response_to_2_800_natural_images/6845348}{figshare}, together with the image stimuli.
\item
  Code availability: The code is available on \href{https://github.com/MouseLand/stringer-pachitariu-et-al-2018b}{GitHub}.
\end{itemize}

\subsubsection{Stringer et al 2019, Science~\citep{stringer_spontaneous_2019}}\label{stringer-et-al-2019-science-stringer2019science}
\begin{itemize}
\item
  Stringer et al.~analyzed spontaneous neural firing, finding that neurons in the primary visual cortex encoded both visual information and motor activity related to facial movements. The variability of neural responses to visual stimuli in the primary visual area is mainly related to arousal and reflects the encoding of latent behavioral states.
\item
  see also the work showing that you can encode very precise orientation information by using many neurons:~\citep{stringer_high-precision_2021}
\end{itemize}

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{detecting structured temporal patterns}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Paper by Grossberger, 2018~\citep{grossberger_unsupervised_2018}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{itemize}
 \item
  Temporally ordered multi-neuron patterns likely encode information in the brain. We introduce an unsupervised method, SPOTDisClust (Spike Pattern Optimal Transport Dissimilarity Clustering), for their detection from high-dimensional neural ensembles. SPOTDisClust measures similarity between two ensemble spike patterns by determining the minimum transport cost of transforming their corresponding normalized cross-correlation matrices into each other (SPOTDis).
\item
  Detecting these temporal patterns represents a major methodological challenge.
\item
  Many approaches to this problem are supervised, that is, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity :
\item
  Nadasdy Z, Hirase H, Czurko A, Csicsvari J, Buzsaki G. Replay and time compression of recurring spike sequences in the hippocampus. J Neurosci. 1999;19(21):9497--507. pmid:10531452
\item
  Lee AK, Wilson MA. A combinatorial method for analyzing sequential firing patterns involving an arbitrary number of neurons based on relative time order. J Neurophysiol. 2004;92(4):2555--73. pmid:15212425
\item
  Davidson TJ, Kloosterman F, Wilson MA. Hippocampal replay of extended experience. Neuron. 2009;63(4):497--507. pmid:19709631
\item
  only one spike per neuron: fig 1A = ``For each pattern and each neuron, a random position was chosen for the activation pulse.''
\item
  t-SNE projection with HDBSCAN labels shows that our clustering method can retrieve all patterns from the data.
\item
  data available @ https://doi.org/10.1371/journal.pcbi.1006283.s013
\end{itemize}

@@

\begin{itemize}
 \item
  ``However, SPOTDis has two principal weaknesses that we address here: (1) Its computational complexity, for comparing two time epochs, is O(N2), where N is the number of neurons. This becomes a major problem for computing an M dissimilarity matrix (for M time epochs) for thousands of neurons. (2) It relies exclusively on pairwise spike-timing relationships (i.e 2nd order correlations), because it does not solve the optimal transport problem for the entire spike pattern, but only for neuron pairs separately. Hence, it may not be sensitive to higher-order correlations in spiking
  patterns.
  Here, we develop a novel dissimilarity measure for multi-neuron spiking patterns called SpikeShip, which has linear computational complexity of O(N). We achieve this by (1) computing the minimum transport cost of spikes for each spike train separately, and (2) discounting a global translation term in the transport flow across neurons.''
  https://doi.org/10.1101/2020.06.03.131573;
\end{itemize}

\subsubsection{Paper by Russo et al 2017~\citep{russo_cell_2017}}\label{paper-by-russo-et-al-2017-russo2017}
\begin{itemize}
 \item
  ``Here we present such a unifying methodological and conceptual framework which detects assembly structure at many different time scales, levels of precision, and with arbitrary internal organization.'' by~\citep{russo_cell_2017}
\item
  sliding window as in~\citep{grun_unitary_2002} (``Numerous other statistical procedures for detecting assemblies or sequential patterns have been proposed previously'') - extended to multiple lags~\citep{torre_synchronous_2016}
\item
  based on a ``non-stationarity-corrected parametric test statistic for assessing the independence of pairs'' and ``an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies''
\end{itemize}

@@

\subsubsection{Cell assemblies at multiple time scales with arbitrary lag constellations}\label{cell-assemblies-at-multiple-time-scales-with-arbitrary-lag-constellations}
~\citep{russo_cell_2017} starts from the relatively old notion of assessing the departure of the joint spike count distribution of two units (or sets) from independence. It is based on unitary events in multiple single-neuron spiking activity~\citep{grun_unitary_2002-1} and the reliable and efficient analysis of an excess or deficiency of joint-spike events~\citep{pipa_neuroxidence_2008}. Having derived a fast, non-stationarity-corrected parametric test statistic for assessing the independence of pairs, they designed an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies.

@@

sparse in time and space : AL Barth and JF Poulet Trends in Neurosciences 35.6 (2012), pp.~345-355. 
CC Petersen and S Crochet, Neuron 78.1 (2013), pp.~28-48.

@@

In~\citep{van_kempen_top-down_2021}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a HMM model for quantifying the transitions. ``In this study, van Kempen et al.~show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives interareal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination.''

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Cortical songs}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{itemize}
 \item
  Ikegaya and colleagues~\citep{ikegaya_synfire_2004} worked on spontaneous activity in vitro and in vivo. They demonstrated that in cortical activity, we can find a repetition of several motifs. In PSCs, but also in spike activity. These sequences repeat after minutes and have a precise spatio temporal structure with a $~\ms$ precision. They can be specific of a particular layer or colomn, are synchronized with network activity oscillation and can involve several cells. They also demonstated that these sequences can form supersequences : the cortical songs. It consist of the assembly of several sequences which repeat in a specific order with a compressed timing.
\item
  ``We find precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons in vivo and in vitro. These patterns repeat after minutes, maintaining millisecond accuracy.''
\item
  Precisely repeating motifs of spontaneous synaptic activity in slices: duration around 1s +/- .5 s. Some events in motifs are of similar size but sometimes absent - better described by Bernouilli than SE (and covariance)
\item
  \emph{in vivo} spontaneous activity also reveals repeating sequences. About 3000 sequences, each involving 3-10 cells out of about 900, and last up to 3 seconds
\item
  topography: ``Sequences had specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster (Fig. 4, A and B, and fig.~S3B). (\ldots) Therefore, repeating temporal patterns of activation (\ldots) were associated with a structured spatial organization of the neurons that formed them.''
\item
  ``Cortical songs: modular assemblies of repeated sequences'': hierarchical detection.
\item
  in cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
\end{itemize}

@@

@@

\subsection{learning pattern detection on natural images / event-based cameras}\label{learning-pattern-detection-on-natural-images-event-based-cameras}
\subsubsection{sparse coding on spatio-temporal data}\label{sparse-coding-on-spatio-temporal-data}
\subsubsection{HOTS}\label{hots}
\subsubsection{Grimaldi CBMI / PAMI}\label{grimaldi-cbmi-pami}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{discussion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spiking neural models with heterogeneous delays}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{our model}\label{our-model}
Here, we develop a model for the efficient detection of such PGs based on the inversion of a probabilistic model defining the generation of the raster plot as a combination of such groups. We show that such an inference can be achieved by a neural-like computation that could itself be used as a spiking neuron, as can be implemented in a neuromorphic chip for instance. A first result is to show the efficiency of such a scheme in detecting different PGs occurring at specific times in synthetic data. The representational capacity of the PGs is particularly interesting compared to traditional models of neuronal encoding using spiking frequency. Our second result is to propose a novel learning method for learning PGs in raster plots in a self-supervised manner. Finally we demonstrate the use of this algorithm to the output of an event-based camera and how this may separate independent components from the stream of events. This end-to-end event-based computational brick could help improve the performance of current Spiking Neural Network solution currently used in neuromorphic chips.

>>>>

%: >>>>>>>> Laurent is here <<<<<<<<<<<<<<
\subsection{Scaling up to very large scale data}
Over the past decade, tremendous technological advances across several disciplines have dramatically expanded the frontiers of experimentally accessible neuroscientific facts. Bridging across different spatial and temporal scales, combination of in vivo two photon imaging, large population recording-array technologies, optogenetic circuit control tools, transgenic manipulations as well as large volume circuit reconstructions are now used to examine the function, structure and dynamics of neural networks on an unprecedented level of detail and precision. The daunting complexity of the biological reality revealed by these technologies highlights the importance of neurophysics to provide a conceptual bridge between abstract principles of brain function and their biological implementations within neural circuits.

One algorithm capable of achieveing such daunting task is the Rastermap algorithm~\citep{pachitariu_robustness_2018}. Basically, it re-arranges neurons in the raster plot based on similarity of activity and applies a deconvolution strategy based on a linear model. The model is \href{https://github.com/MouseLand/rastermap}{openly accessible} and has led to important discoveries. In~\citet{stringer_high-dimensional_2019}, it was shown that a neuronal population encodes information most efficiently when its stimulus responses are high-dimensional and uncorrelated, and most robustly when they are lower-dimensional and correlated. 
Then, in~\citep{stringer_high-precision_2021}, authors analyzed spontaneous neural firing, finding that neurons in the primary visual cortex encoded both visual information and motor activity related to facial movements. Altogether, this provides evidence for the importance of such machine-learning-based tools to provide with breakthroughs in neuroscience.
% \href{https://nbviewer.org/github/MouseLand/rastermap/blob/master/tutorial/tutorial.ipynb}{Rastermap} re-arranges neurons in the raster plot based on similarity of activity

% \begin{itemize}
%  \item
%   https://www.janelia.org/lab/stringer-lab
% \item
%   described in
% \item
%   \href{https://github.com/MouseLand/rastermap}{rastermap}
% \item
%   deconvolution strategy
% \item
%   based on a linear model
% \end{itemize}
% \subsubsection{Stringer et al 2019, Nature~\citep{stringer_high-dimensional_2019}}\label{stringer-et-al-2019-nature-stringer2019nature}
% \begin{itemize}
%  \item
%   ``A neuronal population encodes information most efficiently when its stimulus responses are high-dimensional and uncorrelated, and most robustly when they are lower-dimensional and correlated. Here we analysed the dimensionality of the encoding of natural images by large populations of neurons in the visual cortex of awake mice.''
% \item
%   Data availability: All of the processed deconvolved calcium traces are available on \href{https://figshare.com/articles/Recordings_of_ten_thousand_neurons_in_visual_cortex_in_response_to_2_800_natural_images/6845348}{figshare}, together with the image stimuli.
% \item
%   Code availability: The code is available on \href{https://github.com/MouseLand/stringer-pachitariu-et-al-2018b}{GitHub}.
% \end{itemize}

% \subsubsection{Stringer et al 2019, Science~\citep{stringer_spontaneous_2019}}\label{stringer-et-al-2019-science-stringer2019science}
% \begin{itemize}
% \item
%   Stringer et al.~analyzed spontaneous neural firing, finding that neurons in the primary visual cortex encoded both visual information and motor activity related to facial movements. The variability of neural responses to visual stimuli in the primary visual area is mainly related to arousal and reflects the encoding of latent behavioral states.
% \item
%   see also the work showing that you can encode very precise orientation information by using many neurons:~\citep{stringer_high-precision_2021}
% \end{itemize}

Another important algorithm is based on the detection of structured temporal patterns~\citep{grossberger_unsupervised_2018}. They introduced an unsupervised method based for their detection from high-dimensional neural ensembles. The algorithm measures similarity between two ensemble spike patterns by determining the minimum transport cost of transforming their corresponding normalized cross-correlation matrices into each other. Many approaches to this problem are supervised, that is, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity~\citep{nadasdy_replay_1999,lee_combinatorial_2004}.
% \item
%   Detecting these temporal patterns represents a major methodological challenge.
% \item
%   Many approaches to this problem are supervised, that is, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity :
% \item
%   Nadasdy Z, Hirase H, Czurko A, Csicsvari J, Buzsaki G. Replay and time compression of recurring spike sequences in the hippocampus. J Neurosci. 1999;19(21):9497--507. pmid:10531452
% \item
%   Lee AK, Wilson MA. A combinatorial method for analyzing sequential firing patterns involving an arbitrary number of neurons based on relative time order. J Neurophysiol. 2004;92(4):2555--73. pmid:15212425
% \item
%   Davidson TJ, Kloosterman F, Wilson MA. Hippocampal replay of extended experience. Neuron. 2009;63(4):497--507. pmid:19709631
% \item
In that method, the learning is unsupervised. It uses the prior that there is only one spike per pattern. 
% \item
Using a t-SNE projection with knwon labels validated that this clustering method can retrieve all patterns from the data.
% \item
%   data available @ https://doi.org/10.1371/journal.pcbi.1006283.s013
% \end{itemize}
The limits of this method is that it is block-based and stricly specialized for the task at hand. To overcome these diculties a novel method was recently developped~\citep{sotomayor-gomez_spikeship_2021}.

@@

% \begin{itemize}
%  \item
%   ``However, SPOTDis has two principal weaknesses that we address here: (1) Its computational complexity, for comparing two time epochs, is O(N2), where N is the number of neurons. This becomes a major problem for computing an M dissimilarity matrix (for M time epochs) for thousands of neurons. (2) It relies exclusively on pairwise spike-timing relationships (i.e 2nd order correlations), because it does not solve the optimal transport problem for the entire spike pattern, but only for neuron pairs separately. Hence, it may not be sensitive to higher-order correlations in spiking
%   patterns.
%   Here, we develop a novel dissimilarity measure for multi-neuron spiking patterns called SpikeShip, which has linear computational complexity of O(N). We achieve this by (1) computing the minimum transport cost of spikes for each spike train separately, and (2) discounting a global translation term in the transport flow across neurons.''
%   https://doi.org/10.1101/2020.06.03.131573;
% \end{itemize}

\subsection{Cell assemblies at multiple time scales with arbitrary lag constellations}
\subsubsection{Paper by Russo et al 2017~\citep{russo_cell_2017}}\label{paper-by-russo-et-al-2017-russo2017}
\begin{itemize}
 \item
  ``Here we present such a unifying methodological and conceptual framework which detects assembly structure at many different time scales, levels of precision, and with arbitrary internal organization.'' by~\citep{russo_cell_2017}
\item
  sliding window as in~\citep{grun_unitary_2002} (``Numerous other statistical procedures for detecting assemblies or sequential patterns have been proposed previously'') - extended to multiple lags~\citep{torre_synchronous_2016}
\item
  based on a ``non-stationarity-corrected parametric test statistic for assessing the independence of pairs'' and ``an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies''
\end{itemize}

@@

~\citep{russo_cell_2017} starts from the relatively old notion of assessing the departure of the joint spike count distribution of two units (or sets) from independence. It is based on unitary events in multiple single-neuron spiking activity~\citep{grun_unitary_2002-1} and the reliable and efficient analysis of an excess or deficiency of joint-spike events~\citep{pipa_neuroxidence_2008}. Having derived a fast, non-stationarity-corrected parametric test statistic for assessing the independence of pairs, they designed an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies.

@@

% sparse in time and space : AL Barth and JF Poulet Trends in Neurosciences 35.6 (2012), pp.~345-355. 
CC Petersen and S Crochet, Neuron 78.1 (2013), pp.~28-48.

@@

In~\citep{van_kempen_top-down_2021}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a Hidden Markov Model for quantifying the transitions. In particular, they show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives interareal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination.

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cortical songs}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{itemize}
 \item
  Ikegaya and colleagues~\citep{ikegaya_synfire_2004} worked on spontaneous activity in vitro and in vivo. They demonstrated that in cortical activity, we can find a repetition of several motifs. In PSCs, but also in spike activity. These sequences repeat after minutes and have a precise spatio temporal structure with a $~\ms$ precision. They can be specific of a particular layer or colomn, are synchronized with network activity oscillation and can involve several cells. They also demonstated that these sequences can form supersequences : the cortical songs. It consist of the assembly of several sequences which repeat in a specific order with a compressed timing.
\item
  ``We find precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons in vivo and in vitro. These patterns repeat after minutes, maintaining millisecond accuracy.''
\item
  Precisely repeating motifs of spontaneous synaptic activity in slices: duration around 1s +/- .5 s. Some events in motifs are of similar size but sometimes absent - better described by Bernouilli than SE (and covariance)
\item
  \emph{in vivo} spontaneous activity also reveals repeating sequences. About 3000 sequences, each involving 3-10 cells out of about 900, and last up to 3 seconds
\item
  topography: ``Sequences had specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster (Fig. 4, A and B, and fig.~S3B). (\ldots) Therefore, repeating temporal patterns of activation (\ldots) were associated with a structured spatial organization of the neurons that formed them.''
\item
  ``Cortical songs: modular assemblies of repeated sequences'': hierarchical detection.
\item
  in cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
\end{itemize}

@@

\subsection{learning pattern detection on natural images / event-based cameras}

@@

\subsubsection{sparse coding on spatio-temporal data}\label{sparse-coding-on-spatio-temporal-data}
\subsubsection{HOTS}\label{hots}
\subsubsection{Grimaldi CBMI / PAMI}\label{grimaldi-cbmi-pami}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{discussion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spiking neural models with heterogeneous delays}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{our model}\label{our-model}
Here, we develop a model for the efficient detection of such PGs based on the inversion of a probabilistic model defining the generation of the raster plot as a combination of such groups. We show that such an inference can be achieved by a neural-like computation that could itself be used as a spiking neuron, as can be implemented in a neuromorphic chip for instance. A first result is to show the efficiency of such a scheme in detecting different PGs occurring at specific times in synthetic data. The representational capacity of the PGs is particularly interesting compared to traditional models of neuronal encoding using spiking frequency. Our second result is to propose a novel learning method for learning PGs in raster plots in a self-supervised manner. Finally we demonstrate the use of this algorithm to the output of an event-based camera and how this may separate independent components from the stream of events. This end-to-end event-based computational brick could help improve the performance of current Spiking Neural Network solution currently used in neuromorphic chips.
